---
title: 'Lab 2: Distances and Projections'
subtitle: 'Hints and Tricks'
format:
  html:
    code-fold: true
    toc: true
    include-before-body: ../slides/header.html
    include-after-body:  ../slides/footer-annotations.html
---

```{r, echo = FALSE}
col_dist <- RColorBrewer::brewer.pal(11, "RdGy")
```


```{r, eval = TRUE, message = F, warning = F, echo  = FALSE}
# SPDS
library(tidyverse)
library(sf)
library(units)

# Data
library(USAboundaries)
library(rnaturalearth)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

****

```{r, echo = FALSE, message = FALSE, warning=FALSE}
cities <- read_csv("data/uscities.csv") |> 
  st_as_sf(coords = c("lng", "lat"), crs = 4326) |> 
  st_transform(5070) |> 
  filter(!state_name %in% c("Alaska", "Hawaii", "Puerto Rico"))
```

# Question 1:

### Making Spatial Objects & Coordinate Transformation

Spatial objects (sf) can be built from a vector of X and Y values in addition to a coordinate reference system (CRS). For example:

```{r}
df <- data.frame(name = state.name, 
                X = state.center$x, 
                Y = state.center$y)
head(df)

# Geographic Coordinate System (GCS)
(df_sf_gcs <- st_as_sf(df, 
                      coords = c("X", "Y"), 
                      crs = 4269))

ggplot() + 
  geom_sf(data = df_sf_gcs) + 
  coord_sf(datum = st_crs(df_sf_gcs)) +
  theme_linedraw()

# Projected Coordinate System (PCS)
# st_transforms converts from one reference system to another
(df_sf_pcs = st_transform(df_sf_gcs, 5070))

ggplot() + 
  geom_sf(data = df_sf_pcs) + 
  coord_sf(datum = st_crs(df_sf_pcs)) +
  theme_linedraw()
```

# Question 2: 

### `st_distance` review

```{r}
# Three most populous cities in the USA
(big3 = cities |> 
   select(city, population) |> 
   slice_max(population, n = 3))

# Fort Collins
(fc = filter(cities, city == "Fort Collins") |> 
    select(city, population))

# Distance from FC to population centers
st_distance(big3, fc)
```

There are two notable things about this result:

1. It has `units`
2. It is returned as a matrix, even though Fort Collins only had one row

This second point highlights a useful feature of `st_distance`, namley, its ability to return distance matrices between all combinations of features in `x` and `y`.

### `units` review

While units are useful, they are not always the preferred units. By default, the units measurement is defined by the projection. For example:

```{r, error = FALSE}
st_crs(big3)$units
```

Units can be converted using `units::set_units`. For example, 'm' can be converted to 'km':

```{r}
big3 = mutate(big3, 
              dist_to_fc = st_distance(big3, fc),
              dist_to_fc = set_units(dist_to_fc, "km")) 

(big3$dist_to_fc)
```

You might have noticed  the data `type` of the `st_distance` objects are an S3 class of `units`. Sometimes, this class can cause problems when trying to using it with other classes or methods:

```{r, error = TRUE}
big3$dist_to_fc + 4

ggplot(data = big3) + 
  geom_col(aes(x = city, y = dist_to_fc))
```

In these cases, the units class can be dropped with `units::drop_units`

```{r}
big3 <- mutate(big3, 
              dist_to_fc = st_distance(big3, fc),
              dist_to_fc = set_units(dist_to_fc, "km"),
              dist_to_fc = drop_units(dist_to_fc))

big3$dist_to_fc + 4

ggplot(data = big3) + 
  geom_col(aes(x = reorder(city, -dist_to_fc), y = dist_to_fc)) + 
    labs(title = "Distance to Fort Collins (km)") + 
  ggthemes::theme_fivethirtyeight() + 
  theme( axis.text.x = element_text(face = "bold", size = 14))
```


As with all functions, these steps can be nested:

```{r}
big3 = mutate(big3, 
              dist_to_fc = drop_units(set_units(st_distance(big3, fc), "km")))
```

******

### `Geometry` review

There are a few ways to manipulate existing geometries, here we discuss `st_union()`, `st_combine()` and `st_cast()`

- `st_combine()` returns a single, combined geometry, with **no** resolved boundaries.

- `st_union()` returns a single geometry with resolved boundaries

- `st_cast()` casts one geometry type to another

```{r}
(rockies = USAboundaries::us_states() |> 
  filter(name %in% c('Montana', 'Wyoming', 'Colorado', "New Mexico")) |> 
  select(name, geometry))

plot(rockies['name'], key.pos = 1)


# Combine Geometries
(combined_rk = st_combine(rockies))
plot(combined_rk, col = "red")

# Unioned Geometries
(unioned_rk = st_union(rockies))
plot(unioned_rk, col = "red")

# Combine Geometries
line_rk = st_cast(unioned_rk, "MULTILINESTRING")
plot(line_rk, col = "red")
```

# Question 3:

In this section you will extend your growing `ggplot` skills to handle spatial data using `ggrepl` to label significant features; `gghighlight` to emphasize important criteria; and scaled color/fill to create chloropleth represnetations of variables. Below is some example code to provide an example of these tools in action:

### Get some data (review)

```{r}
# Define a state/region classifier and select the southern states
state.of.interest = data.frame(state = state.name, region = state.region) |> 
  filter(region == "South") |> 
  pull(state)

# Get USA states in the southern region and transform to EPSG:5070
state = USAboundaries::us_states() |> 
  filter(name %in% state.of.interest) |> 
  st_transform(5070)

# Get USA congressional districts in the southern region and transform to EPSG:5070
districts =  USAboundaries::us_congressional() |> 
  filter(state_name %in% state.of.interest) |> 
  st_transform(5070)

# Get the 10 most populous cities in the southern region and transform to EPSG:5070
sub_cities = cities |> 
  filter(state_name %in% state.of.interest) |> 
  slice_max(population, n = 10) |> 
  st_transform(5070)
```


### Map

```{r}
ggplot() + 
  # Add districts with a dashed line (lty = 3), 
  # a color gradient from blue to red based on aland, 
  # and a fill aplha of 0.5
  geom_sf(data = districts, aes(fill = aland), lty = 3, alpha = .5) + 
  scale_fill_gradient(low = 'blue', high = "red") +
  # Highlight (keep blue) only those districts witn a land area > 5e10
  gghighlight(aland > 5e10) +
  # Add the state borders with a thicker line and no fill
  geom_sf(data = state, size = 1, fill = "NA") +
  # Add the cities
  geom_sf(data = sub_cities, size= 2, color = "red") +
  # Add labels to the cities
  ggrepel::geom_label_repel(
    data = sub_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(caption = "Disticts with > 5e10 Area Land are highlighted",
       fill = "Area Land") + 
  ggthemes::theme_map() 
```


